|

### 2019-10-15 analysis/paper/manuscript.Rmd
Address latest comments/edits from Greg and Bob

---
title: |
  Improving the communication and accessibility of stock assessment using interactive visualization tools
author: "Paul M. Regular, Gregory J. Robertson, Robert Rogers, Keith P. Lewis"
date: |
  Fisheries and Oceans Canada, Northwest Atlantic Fisheries Center, 80 East White Hills, St. John's, Newfoundland and Labrador, A1C 5X1, Canada
output: 
  bookdown::word_document2:
    reference_docx: template.docx
bibliography: references.bib
csl: apa.csl
abstract: |
  **Running title**: Interactive stock assessment

@@

  **Data Availability Statement**: Data sharing is not applicable to this article as no new data were created or analyzed in this study.

  **Abstract**: Scientists across many fields are faced with the challenge of synthesizing and communicating information from large and complex data sets. The field of stock assessment is no exception as the volume and variety of the data has grown alongside the computational methods used to integrate them. While this growth in data and model complexity has improved the assessments of many stocks, the process of communicating the results to stakeholders and fisheries managers in a meaningful way has become more daunting. The traditional approach of presenting information across a series of static slides often fails to convey the richness of information available and, as such, important patterns and details are easily overlooked. Here we contend that this problem can be mediated through the effective use of new open source tools for building interactive visualizations. These tools allow a broader audience to conduct detailed explorations of the results, leading to a deeper and collective understanding of both the data and models used to inform stock assessments. As a consequence, the peer review process is more open and accessible and the resulting science advice is improved and widely supported.

@@

Here we aim to demonstrate how interactive visualization tools provide an efficient and effective means of exploring and communicating the ever-expanding array of data inputs and model outputs. We especially highlight the use of interactive visuals inside dynamic documents called "dasbhoards". The concept of using a dashboard was borrowed from the business community where they are frequently used to group a series of interactive visuals and tables to provide at-a-glance views of key performance indicators. We first demonstrate the application of these tools to data that are commonly presented in stock assessments to show how they can simplify the detailed exploration of data from long-term monitoring programs. Second, we focus on the modeling aspect of stock assessment and demonstrate how interactive dasbhoards can be used to explore, diagnose and communicate results from an integrated assessment model. This structure corresponds to two important  steps in the stock assessment process: 1) data exploration, and 2) data modeling. In each case, the dashboards were constructed within the framework of the R programming language [@R] and the RStudio IDE [@Rstudio] using three or more of the packages listed in Table (Table \@ref(tab:pkg)).

@@

## Package name and description
pkg <- c(shiny = 'Makes it incredibly easy to build interactive web applications with R. Automatic "reactive" binding between inputs and outputs and extensive prebuilt widgets make it possible to build beautiful, responsive, and powerful applications with minimal effort [@chang2018a]',
         shinydashboard = "Create dashboards with 'Shiny'. This package provides a theme on top of 'Shiny', making it easy to create attractive dashboards [@chang2018b].",
         rmarkdown = "Convert R Markdown documents into a variety of formats [@allaire2018].",
         flexdashboard = "Format for converting an R Markdown document to a grid oriented dashboard. The dashboard flexibly adapts the size of it's components to the containing web page [@allaire2017].",
         plotly = "Easily translate 'ggplot2' graphs to an interactive web-based version and/or create custom web-based visualizations directly from R [@sievert2018].",
         leaflet = "Create and customize interactive maps using the 'Leaflet' JavaScript library and the 'htmlwidgets' package. These maps can be used directly from the R console, from 'RStudio', in Shiny applications and R [@cheng2018].",
         crosstalk = 'Provides building blocks for allowing HTML widgets to communicate with each other, with Shiny or without (i.e. static .html files) [@cheng2016].')

@@

Fisheries and Oceans Canada (DFO) has been conducting a multi-species stratified-random survey across the Newfoundland and Labrador shelf since the 1970s [@rideout2018]. These data are analyzed using a standard stratified analysis via a locally developed R package called RStrap [for details on methodology see @smith1981]. Both the inputs and outputs from RStrap analyses are very large and, depending on the species, the time-series may include > 40 years and span the majority of the Newfoundland and Labrador shelf. In order for people to quickly and reliably explore these data without iteratively modifying R scripts one species at a time, an application (hereafter called RStrap Explorer) was built using a combination of shiny, crosstalk, flexdashboard, plotly and R markdown (Table \@ref(tab:pkg)). RStrap Explorer started as a way to visualize estimates of biomass and abundance trends of specific species by supplying RStrap output to a flexdashboard file designed to organize and visualize the results. The shiny package was latter applied as it allows RStrap to be run in the background and therefore allows the user to dynamically explore estimates from multiple species in the one session. Both crosstalk and plotly were incorporated to allow the user to interact with the data.

RStrap Explorer contains four primary tabs: 1) "Survey Indices" contains stock level estimates of biomass and abundance (Figure \@ref(fig:biomass)), 2) "Age & Length Distributions" contains length and age frequency plots (Figure \@ref(fig:freq)), 3) "Recruitment" displays recruitment indices, and 4) "Help" provides additional context to the survey data and analysis. The number of potential inputs to RStrap are rather large and most do not need to be modified. Some inputs however, do require modification. The input interface for RStrap Explorer is divided into two parts: basic and advanced inputs. Basic inputs are those like species, season, and survey year(s). Advanced inputs allow the user to define the length or age of recruitment, select whether they would like analyses conducted by length and/or age, and so on. This application was built for data exploration and allow the user to obtain quick and easy visualization of survey data, but is not meant to replace the through data checking and analysis required for formal stock assessment.

@@

Synthesizing data from multiple sources presents a key challenge to stock assessment. Analyses of different data sources were traditionally carried out independently and the summaries or parameters from these analyses were used in the assessment model. This approach, however, is less than ideal because information may be lost and uncertainty may be unaccounted for when we "do statistics on the statistics" [@link1999; @maunder2013]. Such issues have largely been curtailed in contemporary stock assessments thanks to advances in statistical computing that have facilitated the analysis of all available data, in as raw a form as appropriate, in a single integrated analysis [@maunder2013]. Specifically, statistical modeling tools such as JAGS [@plummer2003], AD Model Builder [@fournier2012] and Template Model Builder [@kristensen2015] allow the construction of a joint likelihood for an array of observations to, in theory, extract as much information as possible about the biological and fishery processes. From a computational perspective, analyses of a variety of large data sets has never been easier. However, from a human perspective, contemporary stock assessment biologists are faced with the challenge of understanding and integrating data from multiple sources into a single model and communicating the methods and results to stakeholders and fisheries managers. This challenge was palpable for data-rich case of Northern cod.

@@

We must acknowledge there are costs to adopting the use of interactive tools. The first and obvious one being that staff are required to have both the time and training to effectively develop these tools. In our two examples, the developers (first, second and third authors) were relatively proficient with R and have natural aptitudes for programming, but had their formal training in field-based population ecology. While some learning time was needed (a few weeks), the learning curve was relatively minor because there was no need to learn a new programming language (e.g. JavaScript, CSS, etc.). Moreover, a growing user base means that most programming issues were readily addressed through simple internet searches. We expect the accessibility and quality of these tools to continue to improve as more people in the fisheries community develop and use interactive documents. It is our experience that the upfront investment in time has already paid dividends in terms of efficiencies of delivering products for subsequent stock assessments, and in the quality of the data exploration and understanding. Finally, scientific exploration is rarely a linear process, and the ability to navigate around a dashboard fits the natural way people pursue ideas.

>>>>

---
title: |
  Improving the communication and accessibility of stock assessment using interactive visualization tools
author: "Paul M. Regular, Gregory J. Robertson, Robert Rogers, Keith P. Lewis"
date: |
  Fisheries and Oceans Canada, Northwest Atlantic Fisheries Center, 80 East White Hills, St. John's, Newfoundland and Labrador, A1C 5X1, Canada
output: 
  bookdown::word_document2:
    reference_docx: template.docx
bibliography: references.bib
csl: ices.csl
abstract: |
  **Running title**: Interactive stock assessment

@@

  **Data Availability Statement**: Data sharing is not applicable to this article as no data were generated and no analysis to draw scientific inference was conducted in this study.

  **Corresponding author**: Paul M. Regular, Fisheries and Oceans Canada, Northwest Atlantic Fisheries Center, 80 East White Hills, St. John's, Newfoundland and Labrador, A1C 5X1, Canada, E-mail: Paul.Regular@dfo-mpo.gc.ca, Phone: 1-709-772-2067

  **Abstract**: Scientists across many fields are faced with the challenge of synthesizing and communicating information from large and complex data sets. The field of stock assessment is no exception as the volume and variety of the data has grown alongside the computational methods used to integrate them. While this growth in data and model complexity has improved the assessments of many stocks, the process of communicating the results to colleagues, stakeholders and fisheries managers in a meaningful way has become more daunting. The traditional approach of presenting information across a series of static slides often fails to convey the richness of information available and, as such, important patterns and details are easily overlooked. Here we contend that this problem can be mediated through the effective use of new open source tools for building interactive visualizations. These tools allow a broader audience to conduct detailed explorations of the results, leading to a deeper and collective understanding of both the data and models used to inform stock assessments. As a consequence, the peer review process is more open and accessible and the resulting science advice is improved and widely supported.

@@

Here we aim to demonstrate how interactive visualization tools provide an efficient and effective means of exploring and communicating the ever-expanding array of data inputs and model outputs. We especially highlight the use of interactive visuals inside dynamic documents called "dashboards". The concept of using a dashboard was borrowed from the business community where they are frequently used to group a series of interactive visuals and tables to provide at-a-glance views of key performance indicators. We first demonstrate the application of these tools to data that are commonly presented in stock assessments to show how they can simplify the detailed exploration of data from long-term monitoring programs. Second, we focus on the modeling aspect of stock assessment and demonstrate how interactive dashboards can be used to explore, diagnose and communicate results from an integrated assessment model. This structure corresponds to two important  steps in the stock assessment process: 1) data exploration, and 2) data modeling. In each case, the dashboards were constructed within the framework of the R programming language [@R] and the RStudio IDE [@Rstudio] using three or more of the packages listed in Table (Table \@ref(tab:pkg)).

@@

## Package name and description
pkg <- c(shiny = 'Makes it incredibly easy to build interactive web applications with R. Automatic "reactive" binding between inputs and outputs and extensive prebuilt widgets make it possible to build beautiful, responsive, and powerful applications with minimal effort [@chang2018a]',
         shinydashboard = "Create dashboards with 'Shiny'. This package provides a theme on top of 'Shiny', making it easy to create attractive dashboards [@chang2018b].",
         rmarkdown = "Convert R Markdown documents into a variety of formats [@allaire2018].",
         flexdashboard = "Format for converting an R Markdown document to a grid oriented dashboard. The dashboard flexibly adapts the size of its components to the containing web page [@allaire2017].",
         plotly = "Easily translate 'ggplot2' graphs to an interactive web-based version and/or create custom web-based visualizations directly from R [@sievert2018].",
         leaflet = "Create and customize interactive maps using the 'Leaflet' JavaScript library and the 'htmlwidgets' package. These maps can be used directly from the R console, from 'RStudio', in Shiny applications and R [@cheng2018].",
         crosstalk = 'Provides building blocks for allowing HTML widgets to communicate with each other, with Shiny or without (i.e. static .html files) [@cheng2016].')

@@

Fisheries and Oceans Canada (DFO) has been conducting a multi-species stratified-random survey across the Newfoundland and Labrador shelf since the 1970s [@rideout2018]. These data are analyzed using a standard stratified analysis via a locally developed R package called RStrap [for details on methodology see @smith1981]. Both the inputs and outputs from RStrap analyses are very large and, depending on the species, the time-series may include > 40 years and span the majority of the Newfoundland and Labrador shelf. In order for people to quickly and reliably explore these data without iteratively modifying R scripts one species at a time, an application (hereafter called RStrap Explorer) was built using a combination of shiny, crosstalk, flexdashboard, plotly and R markdown (Table \@ref(tab:pkg)). RStrap Explorer started as a way to visualize estimates of biomass and abundance trends of specific species by supplying RStrap output to a flexdashboard file designed to organize and visualize the results. The shiny package was latter applied as it allows RStrap to be run in the background and therefore allows the user to dynamically explore abundance and biomass estimates from multiple species in one session. Both crosstalk and plotly were incorporated to allow the user to interact with the data.

RStrap Explorer contains four primary tabs: 1) "Survey Indices" contains stock level estimates of biomass and abundance (Figure \@ref(fig:biomass)), 2) "Age & Length Distributions" contains length and age frequency plots (Figure \@ref(fig:freq)), 3) "Recruitment" displays recruitment indices, and 4) "Help" provides additional context to the survey data and analysis. The input interface for RStrap Explorer is divided into two parts: basic and advanced inputs. Basic inputs are those like species, season, and survey year(s). Advanced inputs allow the user to define the length or age of recruitment, select whether they would like analyses conducted by length and/or age, and so on. This application was built for data exploration and allow the user to obtain quick and easy visualization of survey data, before the thorough the data checking and analysis required for formal stock assessment begins.

@@

Synthesizing data from multiple sources presents a key challenge to stock assessment. Analyses of different data sources were traditionally carried out independently and the summaries or parameters from these analyses were used in the assessment model. This approach, however, is less than ideal because information may be lost and uncertainty may be unaccounted for when we "do statistics on the statistics" [@link1999; @maunder2013]. Such issues have largely been curtailed in contemporary stock assessments thanks to advances in statistical computing that have facilitated the analysis of all available data, in as raw a form as appropriate, in a single integrated analysis [@maunder2013]. Specifically, statistical modeling tools such as JAGS [@plummer2003], AD Model Builder [@fournier2012] and Template Model Builder [@kristensen2015] allow the construction of a joint likelihood for an array of observations to, in theory, extract as much information as possible about the biological and fishery processes. From a computational perspective, analyses of a variety of large data sets has never been easier. However, from a human perspective, contemporary stock assessment biologists are faced with the challenge of understanding and integrating data from multiple sources into a single model and communicating the methods and results to stakeholders and fisheries managers. This challenge was palpable for the data-rich case of Northern cod.

@@

We must acknowledge there are costs to adopting the use of interactive tools. The first and obvious one being that staff are required to have both the time and training to effectively develop these tools. In our three examples, the developers (first, second and third authors) were relatively proficient with R and have natural aptitudes for programming, but all had their formal training in field-based population ecology. While some learning time was needed (a few weeks), the learning curve was relatively minor because there was no need to learn a new programming language (e.g. JavaScript, CSS, etc.). Moreover, a growing user base means that most programming issues were readily addressed through simple internet searches. We expect the accessibility and quality of these tools to continue to improve as more people in the fisheries community develop and use interactive documents. It is our experience that the upfront investment in time has already paid dividends in terms of efficiencies of delivering products for subsequent stock assessments, and in the quality of the data exploration and understanding. Finally, scientific exploration is rarely a linear process, and the ability to navigate around a dashboard fits the natural way people pursue ideas.
